\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-12 submission/manuscript.tex
Manuscript 1st version complete.

\vspace{70mm}

\textsuperscript{1\(\dagger\)}

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:marino.korlevic@irb.hr}{\nolinkurl{marino.korlevic@irb.hr}}

1. Ruđer Bošković Institute, Center for Marine Research, G. Paliaga 5,
Rovinj, Croatia

2. University of Vienna, Department of Limnology and Bio-Oceanography,
Althanstraße 14, Vienna, Austria \newpage \linenumbers
\sisetup{mode=text} \setlength\parindent{24pt}

@@

@@

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke \emph{et al.}, 2011b; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal changes
were taken into account low temporal frequency and/or methodologies that
do not allow for high taxonomic resolution were used (Tujula \emph{et
al.}, 2010; Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013;
Michelou \emph{et al.}, 2013). In the present study we describe the
seasonal dynamics of bacterial and archaeal communities on the surfaces
of the seagrass \emph{Cymodocea nodosa} and siphonous macroalgae
\emph{Caulerpa cylindracea} determined on a mostly monthly scale.
Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{Cymodocea nodosa} invaded by the invasive \emph{Caulerpa
cylindracea} and in a locality of only \emph{Caulerpa cylindracea}
located in the proximity of the meadow. In addition, for comparison, the
community of the surrounding seawater was characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the ambient seawater were also processed for
comparison. The number of reads per sample ranged between 8,407 and
77,465 sequences (\autoref{supp-nseq_notus}). Even when the highest
sequencing effort was applied the rarefaction curves did not level off
that is a common observation in high-throughput 16S rRNA amplicon
sequencing approaches (\autoref{supp-rarefaction}). Following quality
curation and exclusion of sequences mentioned before reads were
clustered into 28,726 different OTUs at a similarity level of 97
\si{\percent}. Reads numbers were normalized to the minimum number of
sequences, 8,407 (\autoref{supp-nseq_notus}), through rarefaction
resulting in 17,007 different OTUs that ranged from 366 to 1,998 OTUs
per sample (\autoref{supp-calculators}). To determine seasonal changes
of richness and diversity the Observed Number of OTUs, Chao1, ACE,
Exponential Shannon (Jost, 2006) and Inverse Simpson were calculated
after normalization through rarefaction. Generally, richness estimators
and diversity indices showed similar trends. On average, higher values
were found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,688.5 ± 125.6 OTUs{]} and noninvaded {[}Number of OTUs, 1,744.8 ±
150.8 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,062.8 ± 209.6 OTUs) and lower values for picoplankton
communities in the ambient seawater (Number of OTUs, 528.1 ± 135.8 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{Cymodocea nodosa} communities showed a slow
increase towards the end of the study, while \emph{Caulerpa cylindracea}
(invaded and noninvaded) communities were characterized by slightly
larger values in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were found
at every time point. These OTUs made also a high proportion of total
sequences (39.1 -- 51.3 \si{\percent}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of
\emph{Cymodocea nodosa} associated communities indicated a separation
between Spring, Summer and Autumn/Winter samples (ANOSIM, R = 0.56,
\emph{p} \textless{} 0.01), while communities from the surfaces of
\emph{Caulerpa cylindracea} indicated a non so strongly supported, as in
previous cases, separation between Summer and Autumn/Winter/Spring
samples (ANOSIM, R = 0.31, \emph{p} \textless{} 0.05) (\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by bacterial (99.1 ± 2.1 \si{\percent}) over
archaeal sequences (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{Cymodocea nodosa} were consisted of same groups
with the addition of \emph{Planctomycetota} whose contribution was
higher in summer 2018. In addition, communities from invaded and
noninvaded \emph{Caulerpa cylindracea} were similar and characterized by
same groups as seawater and \emph{Cymodocea nodosa} communities with the
addition of \emph{Desulfobacterota} (\autoref{community}). Larger
differences between environments and host species could be observed at
lower taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{Cymodocea nodosa} (16.4 ± 5.3
\si{\percent}) and \emph{Caulerpa cylindracea} (invaded {[} (7.7 ± 3.9
\si{\percent}){]} and noninvaded {[} (7.8 ± 2.4 \si{\percent}){]})
associated communities in autumn and for seawater communities in winter
(8.8 ± 7.4 \si{\percent}). Large taxonomic differences between surface
associated and seawater cyanobacterial communities were observed.
Seawater communities were mainly comprised of \emph{Cyanobium} and
\emph{Synechococcus}, while surface associated communities were
consisted of \emph{Pleurocapsa} and sequences without known relatives
within \emph{Cyanobacteriia} (\autoref{cyano}). In addition, seasonal
changes in surface associated communities were observed with
\emph{Pleurocapsa} and no relative \emph{Cyanobacteriia} comprising
larger proportions in autumn and winter and \emph{Acrophormium},
\emph{Phormidesmis} and no relative \emph{Nodosilineaceae} in spring and
summer (\autoref{cyano}).

@@

On average, \emph{Alphaproteobacteria} were in comparison to other high
rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to
previous taxa, high differences between seawater and surface associated
communities were observed. Picoplankton communities were composed mainly
of the SAR11 clade, AEGEAN-169 marine group, SAR116 clade, no relative
\emph{Rhodobacteraceae}, HIMB11 and OCS116 clade, while surface
associated communities were composed in high proportion of no relative
\emph{Rhodobacteraceae} and to a lesser degree of \emph{Pseudoahrensia},
no relative \emph{Alphaproteobacteria}, no relative
\emph{Hyphomonadaceae} and \emph{Amylibacter}. Representatives of no
relative \emph{Rhodobacteraceae} were comprising on average 40.6 ± 23.2
\si{\percent} of all alphaproteobacterial sequences from the epiphytic
community (\autoref{alpha}). In addition, \emph{Amylibacter} was
detected mainly in \emph{Cymodocea nodosa} from November 2017 until
March 2018.

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.7 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{Cymodocea nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{Caulerpa cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{Caulerpa cylindracea} specific taxa no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} showed some seasonality and
were characterisitic for samples originating from June to October 2018.
In addition, similarly to \emph{Bacteroidota}, a large proportion of the
surface associated community was grouped as other
\emph{Gammaproteobacteria} indicating high diversity within this group
(\autoref{gamma}).

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{Caulerpa cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{Cymodocea nodosa} communities were consisted of
only 0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{Caulerpa cylindracea} communities their proportion was
25.7 ± 11.3 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Surfaces of marine macrophytes are harboring biofilms consisted of
diverse microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). No standard protocol has been developed to study these
macophyte associated microbes (Ugarelli \emph{et al.}, 2019). Different
procedures of microbial cells removal from host surfaces were described,
such as host tissue shaking (Nõges \emph{et al.}, 2010), scraping (Uku
\emph{et al.}, 2007) and ultrasonication (Cai \emph{et al.}, 2014). All
these methods showed different removal efficiencies but none was
enabling a complete removal of attached microbial cells. In the present
study, we applied an earlier developed removal protocol (Korlević
\emph{et al.}, submitted), based on a previous idea of direct cellular
lysis (Burke \emph{et al.}, 2009), to ensure an almost complete cell
detachment. The application of a direct lysis procedure coupled with a
high frequency sampling protocol and Illumina high resolution amplicon
sequencing has enabled us to make a detailed description of bacterial
and archaeal communities associated with the surfaces of two marine
macrophytes, \emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea}.

In the present study, highest richness values were observed for
\emph{Caulerpa cylindracea} (invaded and noninvaded), middle for
\emph{Cymodocea nodosa} and lowest for seawater derived communities.
Higher values for seagrass associated communities in comparison to
seawater were described earlier and could be attributed to a larger set
of inhabitable microniches existing on macrophyte surfaces (Ugarelli
\emph{et al.}, 2019). In addition, highest values observed for
\emph{Caulerpa cylindracea} are probably a consequence of part of
epiphytic OTUs that are in contact with the sediment. \emph{Caulerpa
cylindracea} stolon is attached to the sediment surface with rhizoids,
so the stolon and rhizoids are in a direct contact with the sediment.
Part of the surface attached \emph{Caulerpa cylnidracea} community is
therefore comprised of OTUs that are in contact with a different
environment and that could cause the observed increase in richness. In
addition, seasonal richness differences observed for surface attached
communities showed slightly higher values in spring and summer. This
pattern could be explained by a higher macrophyte growth in these
seasons (M. Najdek, personal communication; Zavodnik \emph{et al.},
1998; Ruitton \emph{et al.}, 2005). During active periods macrophytes
exhibit a more dynamic chemical interaction with the surface community
probably causing an increase in the number of inhabitable microniches
(Borges and Champenois, 2015; Rickert \emph{et al.}, 2016).

@@

Beside differences between seawater and surface associated communities,
there were discussions if the prokaryotic epiphytic community is
host-specific or there are generalists taxa characteristic to all or
many macrophytes (Egan \emph{et al.}, 2013). Similarly to previously
described differences between seawater and surface attached communities,
at high taxonomic ranks no strong differentiation between communities
associated with different host was observed. The only high rank phylum
that was differing between \emph{Cymodocea nodosa} ans \emph{Caulerpa
cylindracea} was \emph{Desulfobacterota}, whose sequences were more
abundant in the \emph{Caulerpa cylindracea} associated community. As
already mentioned, the rhizoids and part of the stolon are in contact
with the sediment, so \emph{Desulfobacterota} are probably a part of the
epiphytic community that is in contact with the sediment. Similar high
rank taxa found in this study were described to be specific for other
species of macrophytes (Burke \emph{et al.}, 2011b; Lachnit \emph{et
al.}, 2011; Bengtsson \emph{et al.}, 2017). In contrast to high
taxonomic ranks, a substantial differentiation between host specific
communities was found, which supports the host-specific hypothesis.
Similar host-specificity was observed for different species of
macroalgae and seagrasses (Lachnit \emph{et al.}, 2011; Roth-Schulze
\emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019; Morrissey \emph{et
al.}, 2019). Taken together, at high taxonomic ranks a core set of taxa
could be described that is characteristic for all or many macrophytes,
while at low taxonomic ranks a community specific to host species could
be identified (Egan \emph{et al.}, 2013).

>>>>

\vspace{50mm}

Marino Korlević\textsuperscript{1\(*\)}, Marsej
Markovski\textsuperscript{1}, Zihao Zhao\textsuperscript{2}, Gerhard J.
Herndl\textsuperscript{2,3}, Mirjana Najdek\textsuperscript{1}

@@

\textsuperscript{\(*\)}To whom correspondence should be addressed:
\href{mailto:marino.korlevic@irb.hr}{\nolinkurl{marino.korlevic@irb.hr}}

1. Center for Marine Research, Ruđer Bošković Institute, Croatia

2. Department of Functional and Evolutionary Ecology, University of
Vienna, Austria

3. Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on macrophyte
epiphytic communities did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities sampled from the surfaces of the seagrass
\emph{Cymodocea nodosa} and invasive macroalga \emph{Caulerpa
cylindracea}. Leaves and thalli were sampled in a meadow of
\emph{Cymodocea nodosa} invaded by the invasive \emph{Caulerpa
cylindracea} and in a monospecific settlement of \emph{Caulerpa
cylindracea} located in the proximity of the meadow at monthly
intervals. For comparison the ambient prokaryotic plankton community was
also characterized. Sequencing results at the OTU level showed a clear
differentiation between ambient water and epiphytic communities and a
host-specific community assemblage. In addition, successional changes
were observed that could be connected to the macrophyte growth cycle.
Taxonomic analysis showed similar high rank groups in the ambient water
and epiphytic communities, with the exception of \emph{Desulfobacterota}
that were found only on \emph{Caulerpa cylindracea}. Only
\emph{Cyanobacteria} showed seasonal change, while other high rank taxa
were present throughout the year. In every analyzed high rank taxa,
phylogenetic groups present throughout the year and comprising most of
the sequences could be identified together with low proportion taxa
showing seasonal patterns connected to the macrophyte growth cycle.
Taken together, epiphytic microbial communities of the seagrass
\emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa cylindracea}
appear to be host-specific and contain taxa that undergo successional
changes.

@@

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke \emph{et al.}, 2011b; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal changes
were taken into account low temporal frequency and/or methodologies that
do not allow for high taxonomic resolution were used (Tujula \emph{et
al.}, 2010; Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013;
Michelou \emph{et al.}, 2013). In the present study we describe the
seasonal dynamics of bacterial and archaeal communities on the surfaces
of the seagrass \emph{Cymodocea nodosa} and siphonous macroalgae
\emph{Caulerpa cylindracea} determined on a mostly monthly scale.
Bacterial and archaeal epiphytes were sampled in a meadow of \emph{C.
nodosa} invaded by the invasive \emph{C. cylindracea} and in a locality
of only \emph{C. cylindracea} located in the proximity of the meadow. In
addition, for comparison, the community of the ambient seawater was
characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,410 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,729 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,410
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,019
different OTUs that ranged from 355 to 2,018 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (invaded {[}Number of OTUs, 1,690.5 ±
152.3 OTUs{]} and noninvaded {[}Number of OTUs, 1,733.8 ± 163.5
OTUs{]}), middle vales for \emph{C. nodosa} (Number of OTUs, 1,066.9 ±
224.5 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 526 ± 145.6 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (invaded and
noninvaded) communities were characterized by slightly larger values in
Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs made also a high proportion of total sequences (39.9 -- 52.8
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.54, \emph{p} \textless{} 0.01),
while communities from the surfaces of \emph{C. cylindracea} indicated a
non so strongly supported, as in previous cases, separation between
Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31, \emph{p}
\textless{} 0.05) (\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by bacterial (99.1 ± 2.1 \si{\percent}) over
archaeal sequences (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{C. nodosa} were consisted of same groups with the
addition of \emph{Planctomycetota} whose contribution was higher in
summer 2018. In addition, communities from invaded and noninvaded
\emph{C. cylindracea} were similar and characterized by same groups as
seawater and \emph{C. nodosa} communities with the addition of
\emph{Desulfobacterota} (\autoref{community}). Larger differences
between environments and host species could be observed at lower
taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (invaded {[} (7.7 ± 3.9 \si{\percent}){]} and
noninvaded {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.4 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

@@

On average, \emph{Alphaproteobacteria} were in comparison to other high
rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to
previous taxa, high differences between seawater and surface associated
communities were observed. Picoplankton communities were composed mainly
of the SAR11 clade, AEGEAN-169 marine group, SAR116 clade, no relative
\emph{Rhodobacteraceae}, HIMB11 and OCS116 clade, while surface
associated communities were composed in high proportion of no relative
\emph{Rhodobacteraceae} and to a lesser degree of \emph{Pseudoahrensia},
no relative \emph{Alphaproteobacteria}, no relative
\emph{Hyphomonadaceae} and \emph{Amylibacter}. Representatives of no
relative \emph{Rhodobacteraceae} were comprising on average 40.6 ± 23.2
\si{\percent} of all alphaproteobacterial sequences from the epiphytic
community (\autoref{alpha}). In addition, \emph{Amylibacter} was
detected mainly in \emph{C. nodosa} from November 2017 until March 2018.

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.7 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characterisitic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{C. cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{C. nodosa} communities were consisted of only
0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{C. cylindracea} communities their proportion was 25.7 ±
11.2 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Surfaces of marine macrophytes are harboring biofilms consisted of
diverse microbial taxa (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). No standard protocol has been developed to study these
macophyte associated microbes (Ugarelli \emph{et al.}, 2019). Different
procedures of microbial cells removal from host surfaces were described,
such as host tissue shaking (Nõges \emph{et al.}, 2010), scraping (Uku
\emph{et al.}, 2007) and ultrasonication (Cai \emph{et al.}, 2014). All
these methods showed different removal efficiencies but none was
enabling a complete removal of attached microbial cells. In the present
study, we applied an earlier developed removal protocol (Korlević
\emph{et al.}, submitted), based on a previous idea of direct cellular
lysis (Burke \emph{et al.}, 2009), to ensure an almost complete cell
detachment. The application of a direct lysis procedure coupled with a
high frequency sampling protocol and Illumina high resolution amplicon
sequencing has enabled us to make a detailed description of bacterial
and archaeal communities associated with the surfaces of two marine
macrophytes, \emph{C. nodosa} and \emph{C. cylindracea}.

In the present study, highest richness values were observed for \emph{C.
cylindracea} (invaded and noninvaded), middle for \emph{C. nodosa} and
lowest for seawater derived communities. Higher values for seagrass
associated communities in comparison to seawater were described earlier
and could be attributed to a larger set of inhabitable microniches
existing on macrophyte surfaces (Ugarelli \emph{et al.}, 2019). In
addition, highest values observed for \emph{C. cylindracea} are probably
a consequence of part of epiphytic OTUs that are in contact with the
sediment. \emph{C. cylindracea} stolon is attached to the sediment
surface with rhizoids, so the stolon and rhizoids are in a direct
contact with the sediment. Part of the surface attached \emph{Caulerpa
cylnidracea} community is therefore comprised of OTUs that are in
contact with a different environment and that could cause the observed
increase in richness. In addition, seasonal richness differences
observed for surface attached communities showed slightly higher values
in spring and summer. This pattern could be explained by a higher
macrophyte growth in these seasons (M. Najdek, personal communication;
Zavodnik \emph{et al.}, 1998; Ruitton \emph{et al.}, 2005). During
active periods macrophytes exhibit a more dynamic chemical interaction
with the surface community probably causing an increase in the number of
inhabitable microniches (Borges and Champenois, 2015; Rickert \emph{et
al.}, 2016).

@@

Beside differences between seawater and surface associated communities,
there were discussions if the prokaryotic epiphytic community is
host-specific or there are generalists taxa characteristic to all or
many macrophytes (Egan \emph{et al.}, 2013). Similarly to previously
described differences between seawater and surface attached communities,
at high taxonomic ranks no strong differentiation between communities
associated with different host was observed. The only high rank phylum
that was differing between \emph{C. nodosa} ans \emph{C. cylindracea}
was \emph{Desulfobacterota}, whose sequences were more abundant in the
\emph{C. cylindracea} associated community. As already mentioned, the
rhizoids and part of the stolon are in contact with the sediment, so
\emph{Desulfobacterota} are probably a part of the epiphytic community
that is in contact with the sediment. Similar high rank taxa found in
this study were described to be specific for other species of
macrophytes (Burke \emph{et al.}, 2011b; Lachnit \emph{et al.}, 2011;
Bengtsson \emph{et al.}, 2017). In contrast to high taxonomic ranks, a
substantial differentiation between host specific communities was found,
which supports the host-specific hypothesis. Similar host-specificity
was observed for different species of macroalgae and seagrasses (Lachnit
\emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et
al.}, 2019; Morrissey \emph{et al.}, 2019). Taken together, at high
taxonomic ranks a core set of taxa could be described that is
characteristic for all or many macrophytes, while at low taxonomic ranks
a community specific to host species could be identified (Egan \emph{et
al.}, 2013).

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 52.8
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass decay. In contrast, \emph{C.
cylindracea} started to proliferate in late Spring and was characterized
only by a Summer community during maximal biomass increase and by a
Autumn/Winter/Spring community when the biomass were at the peak and the
settlement started to subsequently decay. Similar seasonal changes in
the epiphytic community was described also for other macroalgae (Tujula
\emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). Higher temporal
stability of \emph{C. cylindracea} surface communities in comparison to
\emph{C. nodosa} were also observed in the higher proportion of shared
communities between two consecutive sampling points.

Analysis of seasonal chloroplast sequence abundances showed higher
values in Autumn/Winter in comparison to Spring/Summer. This pattern is
not surprising as e.g seagrasses are known to harbor more epiphytes
during Autumn/Winter (Reyes and Sansón, 2001) and we used an adapted DNA
isolation protocol that is known to partially coextract DNA from
planktonic eukaryotes (Korlević \emph{et al.}, 2015). Strong seasonal
fluctuations of high rank epiphytic taxa were not observed, with the
exception of \emph{Cyanobateria}. Cyanobacterial sequences were more
pronounced in November and December in comparison to Spring and Summer.
